---
title: "Dimensionality Reduction"
author: "Thomas Hollt"
output: github_document
always_allow_html: true
---

# Dimensionality Reduction & Visualization

In this tutorial we will look at different ways to visualize single cell RNA-seq datasets using dimensionality reduction. We will apply the Principal Component Analysis (PCA), t-distributed Stochastic Neighbor Embedding (t-SNE) and Uniform Manifold Approximation and Projection (UMAP) algorithms. Further, we will look at different ways to plot the dimensionality reduced data and augment them with additional information, such as gene expression or meta-information.

## Datasets

For this tutorial we will use the same three human pancreatic islet cell datasets as in the previous tutorial on data integration. The raw data matrix and associated metadata are available [here](https://www.dropbox.com/s/1zxbn92y5du9pu0/pancreas_v3_files.tar.gz?dl=1).

### Data preprocessing

Load required packages. Here, we will use the Seurat functions for the dimensionality reduction. They are also available as pure R functions but Seurat nicely packages them for our data. We will use ggplot for creating the plots.

```{r packages}
suppressMessages(require(Seurat))
suppressMessages(require(ggplot2))
```

### Data loading and preprocessing

Loading and preprocessing is based on the previous data integration lab.

Load in expression matrix and metadata. The metadata file contains the technology (`tech` column) and cell type annotations (`cell type` column) for each cell in the four datasets.
The data is assumed to be in ```~/scrna-seq2019/day3/1-dimensionality-reduction/```. You can just change this to your preferred location, if you do not want to copy the data.

```{r load}
pancreas.data <- readRDS(file = "~/scrna-seq2019/day3/1-dimensionality-reduction/pancreas_expression_matrix.rds")
metadata <- readRDS(file = "~/scrna-seq2019/day3/1-dimensionality-reduction/pancreas_metadata.rds")
```

Create a Seurat object with all datasets as in the previous labs

```{r create_Seurat}
pancreas <- CreateSeuratObject(pancreas.data, meta.data = metadata)
```

We separate the individual datasets into new Seurat objects.

Note: from here on we only use the smartseq2 dataset to manage computation time. You can pick any of the above or the combined. `smartseq2` is the largest of the separated, so if you want to speed up the workflow you can switch to the others. Check the `pancreas.list` for the individual sizes. Make sure you do not remove the objects you want to use.

```{r separate_data}

# split the data by using the tech meta data
pancreas.list <- SplitObject(pancreas, split.by = "tech")

# you can use any subsets from the list. For the lab we continue with the smartseq2 data
# celseq <- pancreas.list[[1]]
# celseq2 <- pancreas.list[[2]]
# fluidigmc1 <- pancreas.list[[3]]
smartseq2 <- pancreas.list[[4]]

# Free up the memory
remove(pancreas.list)
remove(pancreas)
remove(pancreas.data)
gc()
```

We use the same normalization and feature selection as in the previous lab.

```{r normalize}
# Normalize and find variable features
smartseq2 <- NormalizeData(smartseq2, verbose = FALSE)
smartseq2 <- FindVariableFeatures(smartseq2, selection.method = "vst", nfeatures = 2000, verbose = FALSE)
    
# Run the standard workflow for visualization and clustering
smartseq2 <- ScaleData(smartseq2, verbose = FALSE)
```

## Dimensionality Reduction

From here on, we will have a look at the different dimensionality reduction methods and their parameterizations.

### PCA

We start with PCA. Seurat provides the `RunPCA` function, we use `smartseq2` as input data. `npcs` refers to the number of prinicpal components to compute. We set it to `100.` This will take a bit longer to compute but will allow use to explore the differences using different numbers of PCs below.
By assigning the result to our `smartseq2` data object it wil lbe available in the object with the default name `pca`.

```{r dim_red_pca}
smartseq2 <- RunPCA(smartseq2, npcs = 100, verbose = FALSE)
```

We can now plot the first two components using the `DimPlot` function. The first argument here is the Seurat data object `smartseq2`. By providing the `Rreduction = "pca"` argument, `DimPlot` looks in the object for the PCA we created and assigned above.

```{r plot_pca}
DimPlot(smartseq2, reduction = "pca")
```

This plot shows some structure of the data. Every dot is a cell, but we do not know which cells belong to which dot, etc.
We can add meta information by the `group.by` parameter. We use the `celltype` that we have read from the metadata object.

```{r plot_pca_labels}
DimPlot(smartseq2, reduction = "pca", group.by = "celltype")
```

We can see that only a few of the labelled cell-types separate well but many are clumped together on the bottom of the plot.

Let's have a look at the PCs to understand a bit better how PCA separates the data. Using the `DimHeatmap` function, we can plot the expression of the top genes for each PC for a number of cells. We will plot the first six components `dims = 1:6` for 500 random cells `cells = 500`. Each component will produce one heatmap, the cells will be the columns in the heatmap and the top genes for each component the rows.
 
```{r plot_pc_heatmaps}
DimHeatmap(smartseq2, dims = 1:6, cells = 500, balanced = TRUE)
```

As discussed in the lecture, and indicated above, PCA is not optimal for visualization, but can be very helpful in reducing the complexity before applying non-linear dimensionality reduction methods. For that, let's have a look how many PCs actually cover the main vairation.
A very simple, fast-to-compute way is simply looking at the standard deviation per PC. We use the `ElbowPlot`.

```{r plot_pc_std_dev}
ElbowPlot(smartseq2, ndims = 100)
```

As we can see there is a very steep drop in standard deviation within the first 20 or so PCs indicating that we will likelyl be able to use roughly that number of PCs as input to follwing computations with little impact on the results.

### t-SNE

Let's try out t-SNE. Seurat by default uses the [Barnes Hut (BH) SNE implementation](https://arxiv.org/abs/1301.3342).

Similar to the PCA Seurat provies a convenient function to run t-SNE called `RunTSNE`. We provide the `smartseq2` data object as parameter. By default `RunTSNE` will look for and use the PCA we created above as input, we can also force it with `reduction = "pca"`.
Again we use `DimPlot` to plot the result, this time using `reduction = "tsne"` to indicate that we want to plot the t-SNE computation. We create two plots, the first without and the second with the cell-types used for grouping. Already without the coloring, we can see much more structure in the plot than in the PCA plot. With the color overlay we see that most cell-types are nicely separated in the plot.

```{r dim_red_tSNE}
smartseq2 <- RunTSNE(smartseq2, reduction = "pca")

DimPlot(smartseq2, reduction = "tsne")
DimPlot(smartseq2, reduction = "tsne", group.by = "celltype")
```

Above we did not specify the number of PCs to use as input. Let's have a look what happens with different numbers of PCs  as input. We simply run `RunTSNE` multiple times with `dims` defining a range of PCs.
*Note* every run overwrites the `tsne` object nested in the `smartseq2` object. Therefore we plot the tsne directly after each run and store all plots in the `tsne_mult` list. We add `+ NoLegend() + ggtitle("n PCs")` to remove the list of cell types for compactness and add a title.

```{r dim_red_tSNE_multPCA}

# PC_1 to PC_5
smartseq2 <- RunTSNE(smartseq2, reduction = "pca", dims = 1:5)
tsne_mult <- list(DimPlot(smartseq2, reduction = "tsne", group.by = "celltype") + NoLegend() + ggtitle("5 PCs"))

# PC_1 to PC_10
smartseq2 <- RunTSNE(smartseq2, reduction = "pca", dims = 1:10)
tsne_mult[[2]] <- DimPlot(smartseq2, reduction = "tsne", group.by = "celltype") + NoLegend() + ggtitle("10 PCs")

# PC_1 to PC_30
smartseq2 <- RunTSNE(smartseq2, reduction = "pca", dims = 1:30)
tsne_mult[[3]] <- DimPlot(smartseq2, reduction = "tsne", group.by = "celltype") + NoLegend() + ggtitle("30 PCs")

# PC_1 to PC_100
smartseq2 <- RunTSNE(smartseq2, reduction = "pca", dims = 1:100)
tsne_mult[[4]] <- DimPlot(smartseq2, reduction = "tsne", group.by = "celltype") + NoLegend() + ggtitle("100 PCs")

CombinePlots(plots = tsne_mult)
```

Looking at these plots, it seems `RunTSNE` by default only uses 5 PCs (the plot is identical to the plot with default parameters above), but we get much clearer separation of clusters using 10 or even 30 PCs. This is not surprising considering the plot above of the standard deviation within the PCs above. Therefore we will use 30 PCs in the following, by explicitly setting `dims = 1:30`. 
*Note* that t-SNE is slower with more input dimensions (here the PCs), so it is good to find a middleground between capturing as much variation as possible with as few PCs as possible. However, using the default 5 does clearly not produce optimal results for this dataset.
When using t-SNE with PCA preprocessing with your own data, always check how many PCs you need to cover the variance, as done above.


t-SNE has a few hyper-parameters that can be tuned for better visualization. There is an [excellent tutorial](https://distill.pub/2016/misread-tsne/).
The main parameter is the perplexity, basically indicating how many neighbors to look at.
We will run different perplexities to see the effect. As we will see, a perplexity of 30 is the default. This value often works well, again it might be advisable to test different values with other data.
*Note*, higher perplexity values make t-SNE slower to compute.

```{r dim_red_tSNE_perplexity}

# Perplexity 3
smartseq2 <- RunTSNE(smartseq2, reduction = "pca", dims = 1:30, perplexity = 3)
tsne_mult <- list(DimPlot(smartseq2, reduction = "tsne", group.by = "celltype") + NoLegend() + ggtitle("30PCs, Perplexity 3"))

# Perplexity 10
smartseq2 <- RunTSNE(smartseq2, reduction = "pca", dims = 1:30, perplexity = 10)
tsne_mult[[2]] <- DimPlot(smartseq2, reduction = "tsne", group.by = "celltype") + NoLegend() + ggtitle("30PCs, Perplexity 10")

# Perplexity 30
smartseq2 <- RunTSNE(smartseq2, reduction = "pca", dims = 1:30, perplexity = 30)
tsne_mult[[3]] <- DimPlot(smartseq2, reduction = "tsne", group.by = "celltype") + NoLegend() + ggtitle("30PCs, Perplexity 30")

# Perplexity 200
smartseq2 <- RunTSNE(smartseq2, reduction = "pca", dims = 1:30, perplexity = 200)
tsne_mult[[4]] <- DimPlot(smartseq2, reduction = "tsne", group.by = "celltype") + NoLegend() + ggtitle("30PCs, Perplexity 200")

CombinePlots(plots = tsne_mult)
```

Another important parameter is the number of iterations. t-SNE gradually optimizes the low-dimensional space. The more iterations the more there is to optimize.
We will run different numbers of iterations to see the effect. As we will see, 1000 iterations is the default. This value often works well, again it might be advisable to test different values with other data. Especially for larger datasets you will need more iterations.
*Note*, more iterations make t-SNE slower to compute.

```{r dim_red_tSNE_iterations}

# 100 iterations
smartseq2 <- RunTSNE(smartseq2, reduction = "pca", dims = 1:30, max_iter = 100)
tsne_mult <- list(DimPlot(smartseq2, reduction = "tsne", group.by = "celltype") + NoLegend() + ggtitle("100 iterations"))

# 500 iterations
smartseq2 <- RunTSNE(smartseq2, reduction = "pca", dims = 1:30, max_iter = 500)
tsne_mult[[2]] <- DimPlot(smartseq2, reduction = "tsne", group.by = "celltype") + NoLegend() + ggtitle("500 iterations")

# 1000 iterations
smartseq2 <- RunTSNE(smartseq2, reduction = "pca", dims = 1:30, max_iter = 1000)
tsne_mult[[3]] <- DimPlot(smartseq2, reduction = "tsne", group.by = "celltype") + NoLegend() + ggtitle("1000 iterations")

# 2000 iterations
smartseq2 <- RunTSNE(smartseq2, reduction = "pca", dims = 1:30, max_iter = 2000)
tsne_mult[[4]] <- DimPlot(smartseq2, reduction = "tsne", group.by = "celltype") + NoLegend() + ggtitle("2000 iterations")

CombinePlots(plots = tsne_mult)
```

As we see after 100 iterations the main structure becomes apparent, but there is very little detail. 500 to 2000 iterations all look very similar, with 500 still a bit more loose than 1000, indicating that the optimization converges somewhere between 500 and 1000 iterations. In this case running 2000 would definitely not necessary.

### UMAP

We have seen t-SNE and it's main parameters. Let's have a look at UMAP. Its main function call is very similar to t-SNE and PCA and called `RunUMAP`. Again, by default it looks for the PCA in the `smartseq2` data object, but we have to provide it with the number of PCs (or `dims`) to use. Here, we use `30`. As expected, the plot looks rather similar to the t-SNE plot, with more compact clusters.

```{r dim_red_UMAP}
smartseq2 <- RunUMAP(smartseq2, dims = 1:30)

DimPlot(smartseq2, reduction = "umap", group.by = "celltype")
```

Just like t-SNE UMAP has a bunch of parameters. In fact, Seurat exposes wuite a few more than for t-SNE. We will look at the most important in the following. While there is no site that systematically and interactively explores them as for t-SNE there is a [comparison with the same datasets available](https://jlmelville.github.io/uwot/umap-simple.html).

Again, we start with a different number of PCs. similar to t-SNE, 5 is clearly not enough, 30 provides decent separation and detail.
Just like for t-SNE, test this parameter to match your own data in real-world experiments.

```{r dim_red_UMAP_multPCA}

# PC_1 to PC_5
smartseq2 <- RunUMAP(smartseq2, dims = 1:5)
umap_mult <- list(DimPlot(smartseq2, reduction = "umap", group.by = "celltype") + NoLegend() + ggtitle("5 PCs"))

# PC_1 to PC_10
smartseq2 <- RunUMAP(smartseq2, dims = 1:10)
umap_mult[[2]] <- DimPlot(smartseq2, reduction = "umap", group.by = "celltype") + NoLegend() + ggtitle("10 PCs")

# PC_1 to PC_30
smartseq2 <- RunUMAP(smartseq2, dims = 1:30)
umap_mult[[3]] <- DimPlot(smartseq2, reduction = "umap", group.by = "celltype") + NoLegend() + ggtitle("30 PCs")

# PC_1 to PC_100
smartseq2 <- RunUMAP(smartseq2, dims = 1:100)
umap_mult[[4]] <- DimPlot(smartseq2, reduction = "umap", group.by = "celltype") + NoLegend() + ggtitle("100 PCs")

CombinePlots(plots = umap_mult)
```

The `n.neighbors` parameter sets the number of neighbors to consider for UMAP. This parameter is similar to the perplexity in t-SNE.
We try a similar range of values for comparison. The results are quite similar to t-SNE. with low values, clearly the structures are too spread out, but quickly the embeddings become quite stable. The default vaule for RunUMAP is 30. Similar to t-SNE this value is quite general.
Again, it's always a good idea to run some test with new data to find a good value.

```{r dim_red_UMAP_multNeighbors}

# 3 Neighbors
smartseq2 <- RunUMAP(smartseq2, dims = 1:30, n.neighbors = 3)
umap_mult <- list(DimPlot(smartseq2, reduction = "umap", group.by = "celltype") + NoLegend() + ggtitle("3 Neighbors"))

# 10 Neighbors
smartseq2 <- RunUMAP(smartseq2, dims = 1:30, n.neighbors = 10)
umap_mult[[2]] <- DimPlot(smartseq2, reduction = "umap", group.by = "celltype") + NoLegend() + ggtitle("10 Neighbors")

# 30 Neighbors
smartseq2 <- RunUMAP(smartseq2, dims = 1:30, n.neighbors = 30)
umap_mult[[3]] <- DimPlot(smartseq2, reduction = "umap", group.by = "celltype") + NoLegend() + ggtitle("30 Neighbors")

# 200 Neighbors
smartseq2 <- RunUMAP(smartseq2, dims = 1:30, n.neighbors = 200)
umap_mult[[4]] <- DimPlot(smartseq2, reduction = "umap", group.by = "celltype") + NoLegend() + ggtitle("200 Neighbors")

CombinePlots(plots = umap_mult)
```

Another parameter is `min.dist`. There is no directly comparable parameter in t-SNE. Generally, `min.dist` defines the compactnes of the final embedding. The default value is 0.3.

```{r dim_red_UMAP_multMinDist}

# Min Distance 0.01
smartseq2 <- RunUMAP(smartseq2, dims = 1:30, min.dist = 0.01)
umap_mult <- list(DimPlot(smartseq2, reduction = "umap", group.by = "celltype") + NoLegend() + ggtitle("Min Dist 0.01"))

# Min Distance 0.1
smartseq2 <- RunUMAP(smartseq2, dims = 1:30, min.dist = 0.1)
umap_mult[[2]] <- DimPlot(smartseq2, reduction = "umap", group.by = "celltype") + NoLegend() + ggtitle("Min Dist 0.1")

# Min Distance 0.3
smartseq2 <- RunUMAP(smartseq2, dims = 1:30, min.dist = 0.3)
umap_mult[[3]] <- DimPlot(smartseq2, reduction = "umap", group.by = "celltype") + NoLegend() + ggtitle("Min Dist 0.3")

# Min Distance 1.0
smartseq2 <- RunUMAP(smartseq2, dims = 1:30, min.dist = 1.0)
umap_mult[[4]] <- DimPlot(smartseq2, reduction = "umap", group.by = "celltype") + NoLegend() + ggtitle("Min Dist 1.0")

CombinePlots(plots = umap_mult)
```

n.epochs is comparable to the number of iterations in t-SNE. Typically, UMAP needs fewer of these to converge, but also changes more when it is run longer. The default is 500. Again, this should be adjusted to your data.

```{r dim_red_UMAP_multEpochs}

# 10 epochs
smartseq2 <- RunUMAP(smartseq2, dims = 1:30, n.epochs = 10)
umap_mult <- list(DimPlot(smartseq2, reduction = "umap", group.by = "celltype") + NoLegend() + ggtitle("10 Epochs"))

# 100 epochs
smartseq2 <- RunUMAP(smartseq2, dims = 1:30, n.epochs = 100)
umap_mult[[2]] <- DimPlot(smartseq2, reduction = "umap", group.by = "celltype") + NoLegend() + ggtitle("100 Epochs")

# 500 epochs
smartseq2 <- RunUMAP(smartseq2, dims = 1:30, n.epochs = 500)
umap_mult[[3]] <- DimPlot(smartseq2, reduction = "umap", group.by = "celltype") + NoLegend() + ggtitle("500 Epochs")

# 1000 epochs
smartseq2 <- RunUMAP(smartseq2, dims = 1:30, n.epochs = 1000)
umap_mult[[4]] <- DimPlot(smartseq2, reduction = "umap", group.by = "celltype") + NoLegend() + ggtitle("1000 Epochs")

CombinePlots(plots = umap_mult)
```

Finally `RunUMAP` allows to set the distance metric for the high-dimensional space. While in principle this is also possible with t-SNE, `RunTSNE`, and most other implementations, do not provide this option. The four posibilites, `Euclidean`, `cosine`, `manhattan`, and `hamming` distance are shown below. `Cosine` distance is the default (RunTSNE uses Euclidean distances).

There is not necesarily a clear winner. Hamming distances perfom worse but they are usually used for different data, such as text as they ignore the numerical difference for a given comparison. Going with the dafualt cosine is definitely not a bad choice in most applications.

```{r dim_red_UMAP_multMetric}

# Euclidena distance
smartseq2 <- RunUMAP(smartseq2, dims = 1:30, metric = "euclidean")
umap_mult <- list(DimPlot(smartseq2, reduction = "umap", group.by = "celltype") + NoLegend() + ggtitle("Euclidean"))

# Cosine distance
smartseq2 <- RunUMAP(smartseq2, dims = 1:30, metric = "cosine")
umap_mult[[2]] <- DimPlot(smartseq2, reduction = "umap", group.by = "celltype") + NoLegend() + ggtitle("Cosine")

# Manhattan distance
smartseq2 <- RunUMAP(smartseq2, dims = 1:30, metric = "manhattan")
umap_mult[[3]] <- DimPlot(smartseq2, reduction = "umap", group.by = "celltype") + NoLegend() + ggtitle("Manhattan")

# Hamming distance
smartseq2 <- RunUMAP(smartseq2, dims = 1:30, metric = "hamming")
umap_mult[[4]] <- DimPlot(smartseq2, reduction = "umap", group.by = "celltype") + NoLegend() + ggtitle("Hamming")

CombinePlots(plots = umap_mult)
```

## Visualization

Finally, let's have a brief look at some visualization options. We have already use color gor gouping. With `label = TRUE` we can add a text-label to each group and with `repel = TRUE` we can make sure those labels dont clump together. Finally, `pt.size = 0.5` changes the size of the dots used in the plot.

```{r dim_red_vis_prep}
#Re-run a t-SNE so we do not rely on changes above
smartseq2 <- RunTSNE(smartseq2, dims = 1:30)

DimPlot(smartseq2, reduction = "tsne", group.by = "celltype", label = TRUE, repel = TRUE, pt.size = 0.5) + NoLegend()
```

Another property we might want to look at in our dimensionality reduction plot is the expression of individual Genes. We can overlay gene expression as color using the `FeaturePlot`.
Here, we first find the *top two* features correlated to the *first and second PCs* and combine them into a single vector which will be the parameter for the `FeaturePlot`.

Finally, we call `FeaturePlot` with the `smartseq2` data object, `features = topFeaturesPC` uses the extracted feature vector to create one plot for each feature in the list and lastly, `reduction = "pca"` will create PCA plots.

Not surprisingly, the top two features of the first PC form a smooth gradient on the PC_1 axis and the top two features of the second PC a smooth gradient on the PC_2 axis.

```{r dim_red_vis_features_pca}

# find top genes for PCs 1 and 2
topFeaturesPC1 <- TopFeatures(object = smartseq2[["pca"]], nfeatures = 2, dim = 1)
topFeaturesPC2 <- TopFeatures(object = smartseq2[["pca"]], nfeatures = 2, dim = 2)

# combine the genes into a single vector
topFeaturesPC <- c(topFeaturesPC1, topFeaturesPC2)
print(topFeaturesPC)

# feature plot with the defined genes
FeaturePlot(smartseq2, features = topFeaturesPC, reduction = "pca")
```

Let's have a look at the same features on a t-SNE plot. The bahvior here is quite different, with the high expression being very localized to specific clusters in the maps. Again, not surprising as t-SNE uses these 

```{r dim_red_vis_features_tsne}
FeaturePlot(smartseq2, features = topFeaturesPC, reduction = "tsne")
```

It is clear that the top PCs are fundamental to forming the clusters, so let's have a look at more PCs and pick the top gene per PC for a few more PCs.

```{r dim_red_vis_features_tsne2}
for(i in 1:6) {
  topFeaturesPC[[i]] <- TopFeatures(object = smartseq2[["pca"]], nfeatures = 1, dim = i)
}
print(topFeaturesPC)
FeaturePlot(smartseq2, features = topFeaturesPC, reduction = "tsne")
```

Finally, let's create a more interactive plot. First we create a regular `FeaturePlot`, here with just one gene `features = "TM4SF4"`. Instead of plotting in directly we save the plot in the `interactivePlot` variable.

With `HoverLocator` we can then embedd this plot into an interactive version. The `information = FetchData(smartseq2, vars = c("celltype", topFeaturesPC))` creates a set of properties that will be shown on hover over each point. Im this case, we show the cell-type from the meta information.

The result is a plot that allows us to inspect single cells in detail.

```{r dim_red_vis_interactive}
interactivePlot <- FeaturePlot(smartseq2, reduction = "tsne", features = "TM4SF4")

HoverLocator(plot = interactivePlot, information = FetchData(smartseq2, vars = c("celltype", topFeaturesPC)))
```


